---
title: The Relationship Between Geochemical Soil Properties, Parent Rock & Major Geological Events and Processes
author: "500494215"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document:
    code_folding: hide 
    toc: true 
    toc_float: true
    number_sections: true
    df_print: kable
    self_contained: true 

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message = FALSE)
library(sf)
library(mapview)
library(tidyverse)
library(RColorBrewer)

```

```{r include=FALSE}
#Data loading + NA value defining 
data= read_csv("Cooma_pXRF_data.csv", na=c('NO DATA','<LOD'))
chem_data= data %>% select(1:14)
#Drop NA values
na_chem_data=drop_na(chem_data)
#Jitter same coordinates for greater ease to view values in map
data_jitter= data %>% 
  st_as_sf(coords = c("Lng", "Lat"), crs = 4326) %>% 
  st_jitter(factor = 0.02)
```
## Geological History of the Southern Tablelands
- 500 words
- Major events that produced and modified lithology + topography
- use field trip + reading lit

Plateau- 1000m above sea level, east coast is a drop- escarpment 
mostly ancient highland, basaltic bedrock near cooma produce true chernozems in aus- fertile
granitic- infertile and leached
aureole 

## Rock Geochemistry and Surface Geology
- specific geochemical differences between rock types
- VFT + reading- relate geochemical fingerprint of rock types to processes that formed them
- 100 words

## Soil Geochemistry and Soil Types
- Geochemical differences between soil types? What are they?
- rock samples + VFT. Explain why the geochemical fingerprints of soil types are different. 

```{r}
mapviewOptions(basemaps = c("OpenTopoMap", "Esri.WorldShadedRelief", "OpenStreetMap.DE", "CartoDB.Positron"))

mapview(data_jitter, xcol='Lng', ycol='Lat', crs = 4326, grid = FALSE, label= 'Material', zcol ="Si",legend=TRUE)+ mapview(data_jitter, xcol='Lng', ycol='Lat', crs = 4326, grid = FALSE, label= 'Material', zcol = "Al", at = seq(2, 14, 2), legend=TRUE, hide= TRUE, col.regions=brewer.pal(9, "YlOrRd"))

```


```{r}
ggplot(data= na_chem_data, aes(x=Material, y=Al, fill=Location)) + geom_bar(stat = 'identity') + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


## References
https://map-rfun.library.duke.edu/01_georeference.html 
https://r-spatial.github.io/mapview/reference/mapView.html 
https://rdrr.io/github/r-spatial/mapview/f/vignettes/articles/mapview_02-advanced.Rmd 

http://www.regional.org.au/au/asssi/supersoil2004/s12/oral/1777_jenkinsb.htm 

https://www.sciencedirect.com/science/article/pii/S1040618201000386 - landscape evolution, uplift, erosion
